Model-based pretreatment for regression with correlated variables.

### 2014-04-16 www/article13.tex
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@538 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

 Instead of reducing the dimension without looking at correlations,  we propose to focus on the correlations, giving an explicit structure between covariates with the idea that if we know explicitly the correlations, we could use this knowledge to avoid the problem. Practically speaking we use a system of linear regressions for the joint distribution of $\boldsymbol{X}$ to model the structure of the correlations. It helps to define the greatest set of orthogonal covariates to keep the maximum information but with an orthogonality constraint.
Then we can define a marginal model with independent covariates. Thus we have variable selection with explanation and improve interpretation and not only prediction.
 This can be viewed as a pretreatment on the dataset that will be followed then by any other tools for estimation and dimension reduction without suffering from correlations. This pretreatment is specifically done to decorrelate the covariates, unlike methods described above.

@@

\subsection{Our proposal: modelisation of the correlations}
We make the hypothesis that $\boldsymbol{X}$ can be described by a partition $\boldsymbol{X}=(\boldsymbol{X}_f,\boldsymbol{X}_r) $ where variables in $\boldsymbol{X}_r$ are endogenous covariates resulting from linear sub-regressions based on $\boldsymbol{X}_f$, the submatrix of mutually independent exogenous covariates. Then $\boldsymbol{X}_r$ is the $n\times p_r$ submatrix of $0\leq p_r <p$ redundent covariates and $\boldsymbol{X}_f$ the $n\times (p-p_r)$ submatrix of the free (independent) covariates.
In the following, we note $\boldsymbol{X}^j$ the $j^{th}$ column of $\boldsymbol{X}$ and $|.|$ represents the cardinal of an ensemble.
The family of $p_r$ regressions within correlated covariates in $\boldsymbol{X}$ is noted:
	\begin{equation}
		\boldsymbol{X}_{r|\boldsymbol{X}_f,S} \textrm{ defined by }\forall \boldsymbol{X}^j \subset \boldsymbol{X}_r: \boldsymbol{X}^j_{|\boldsymbol{X}_f,S}=\boldsymbol{X}_f\boldsymbol{\alpha}_j+\boldsymbol{\varepsilon}_j \textrm{ with } \boldsymbol{\varepsilon}_j \sim\mathcal{N}(\boldsymbol{0},\sigma^2_j\boldsymbol{I}_n) \label{SR}
	\end{equation}
		where $\boldsymbol{\alpha}_j \in \mathcal{R}^{(p-p_r)}$ are the vectors of the regression coefficients between the covariates. So we now have the expression of $P(\boldsymbol{X}_r|\boldsymbol{X}_f) $
\\
\\
We note $I_r$ the set of indices of endogenous variables in $\boldsymbol{X}$ (explained ones).
We also define $I_f=\{I_f^1,\dots,I_f^p \}$ the set of the sets of indices of exogenous covariates (explaining ones $=\boldsymbol{X}_f$) with $\forall j \notin I_r, I_f^j=\emptyset$. 
We see that $I_f$ defines the non-null coefficients in $\boldsymbol{\alpha}_j$ (each sub-regression can be very parsimonious).
Then we have an explicit structure $S=(I_f,I_r,p_f,p_r)$ where $p_r=|I_r|$, $\boldsymbol{p}_f=(p_f^1,\dots,p_f^{p_r})$ is the vector of the number of covariates in each sub-regression  and $p_f^j=|I_f^j|$. 
\\
\\

@@

\subsection{A by-product model: marginal regression with decorrelated covariates}
Now we know $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)$ by the structure of sub-regressions, we are able to define a marginal regression model $P(\boldsymbol{Y}|\boldsymbol{X}_f,S)$ based on the reduced set of independent covariates $\boldsymbol{X}_f$ without significant information loss. 
 	\\
Using the partition $\boldsymbol{X}=[\boldsymbol{X}_f,\boldsymbol{X}_r]$ we can rewrite (\ref{regressionsimple}):
	\begin{equation}
			\boldsymbol{Y}_{|\boldsymbol{X}_f,\boldsymbol{X}_r,S}=\boldsymbol{X}_f\boldsymbol{\beta}_f+\boldsymbol{X}_r\boldsymbol{\beta}_r+\boldsymbol{\varepsilon_Y} \label{MainR}
		\end{equation}
		where $\boldsymbol{\beta}=(\boldsymbol{\beta}_f,\boldsymbol{\beta}_r) \in  \mathcal{R}^p$ is the vector of the regression coefficients associated respectively to $\boldsymbol{X}_f$ and $\boldsymbol{I}_n$ the identity matrix. 
We note that (\ref{SR}) and (\ref{MainR}) give also by simple integration on $\boldsymbol{X}_r$ a marginal regression model on $\boldsymbol{Y}$ {\it depending only on uncorrelated covariates $\boldsymbol{X}_f$}:
\begin{eqnarray}
	\boldsymbol{Y}_{|\boldsymbol{X}_f,S}&=&\boldsymbol{X}_f (\boldsymbol{\beta}_f+ \sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j)+  \sum_{j \in I_r}\beta_{j}\boldsymbol{\varepsilon}_j+\boldsymbol{\varepsilon}_Y \label{Trueexpl} \\
	&=&\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{\varepsilon}_Y^*=\boldsymbol{X}\boldsymbol{\beta}^*+\boldsymbol{\varepsilon}_Y^* \textrm{ where }\boldsymbol{\beta}^*=(\boldsymbol{\beta}_f^*,\boldsymbol{\beta}_r^*) \textrm{ and } \hat{\boldsymbol{\beta}}_r^*=\boldsymbol{\beta}_r^*=\boldsymbol{0}\label{modexpl}
\end{eqnarray}
We note that it is simply a linear regression on some of the original covariates so we only made a pretreatment on the dataset by setting some coefficients to $0$ because of correlations. These $0$ won't be interpreted as independence with the response variable but as redundant information due to correlations between the covariates. It is a variable pre-selection independent of the response $\boldsymbol{Y}$.

@@

	 The Mean Squared Error (\textsc{MSE}) on $\hat{\boldsymbol{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\hat{\boldsymbol{\beta}}|\boldsymbol{X})&=&\parallel \operatorname{Bias}\parallel_2^2+\operatorname{Tr}(\operatorname{Var}(\hat{\boldsymbol{\beta}})) \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}_{OLS}|\boldsymbol{X})&=& 0 + \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) \textrm{ for OLS, and then for the marginal model:} \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}^*_{OLS}|\boldsymbol{X})&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1})
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $\boldsymbol{X}_f$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $\boldsymbol{x}_3=\boldsymbol{x}_1+\boldsymbol{x}_2+\boldsymbol{\varepsilon}_3$ where $\boldsymbol{\varepsilon}_3\sim{\mathcal{N}(\boldsymbol{0},\sigma_3^2\boldsymbol{I}_n)}$. We also have $\boldsymbol{\beta}=(1,1,1)$ and $\sigma_Y \in \{1,10\}$  . Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg}'s marginal  model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figure \ref{MQE1} shows the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\boldsymbol{\varepsilon)_3}}{\operatorname{Var}(\boldsymbol{x}_3)}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}

>>>>

 Instead of reducing the dimension without looking at correlations,  we propose to focus on the correlations, giving an explicit structure between covariates with the idea that if we know explicitly the correlations, we could use this knowledge to avoid the problem. More precisely, we model the correlations with a system of linear regressions for the joint distribution of $\boldsymbol{X}$ to model the structure of the correlations. It helps to define the greatest set of orthogonal covariates to keep the maximum information but with an orthogonality constraint.
Then we can define a marginal model with independent covariates. Thus we have variable selection with explanation and improve interpretation and not only prediction.
 This can be viewed as a pretreatment on the dataset that will be followed then by any other tools for estimation and dimension reduction without suffering from correlations. This pretreatment is specifically done to decorrelate the covariates, unlike methods described above.

@@

\subsection{Our proposal: modelisation of the correlations}
We make the hypothesis that $\boldsymbol{X}$ can be described by a partition $\boldsymbol{X}=(\boldsymbol{X}_f,\boldsymbol{X}_r) $ where variables in $\boldsymbol{X}_r$ are endogenous covariates resulting from linear sub-regressions based on $\boldsymbol{X}_f$, the submatrix of mutually independent exogenous covariates. Then $\boldsymbol{X}_r$ is the $n\times p_r$ submatrix of $0\leq p_r <p$ redundent covariates and $\boldsymbol{X}_f$ the $n\times (p-p_r)$ submatrix of the free (independent) covariates.
In the following, we note $\boldsymbol{X}^j$ the $j^{th}$ column of $\boldsymbol{X}$ and $|.|$ represents the cardinal of an ensemble.
The structure $S$ of $p_r$ regressions within correlated covariates in $\boldsymbol{X}$ is described by:
	\begin{equation}
		\boldsymbol{X}_{r|\boldsymbol{X}_f,S} \textrm{ defined by }\forall \boldsymbol{X}^j \subset \boldsymbol{X}_r: \boldsymbol{X}^j_{|\boldsymbol{X}_f,S}=\boldsymbol{X}_f\boldsymbol{\alpha}_j+\boldsymbol{\varepsilon}_j \textrm{ with } \boldsymbol{\varepsilon}_j \sim\mathcal{N}(\boldsymbol{0},\sigma^2_j\boldsymbol{I}_n) \label{SR}
	\end{equation}
		where $\boldsymbol{\alpha}_j \in \mathcal{R}^{(p-p_r)}$ are the vectors of the regression coefficients between the covariates. So we now have the expression of $P(\boldsymbol{X}_r|\boldsymbol{X}_f) $
\\
\\
We note $I_r$ the set of indices of endogenous variables in $\boldsymbol{X}$ (explained ones).
We also define $I_f=\{I_f^1,\dots,I_f^p \}$ the set of the sets of indices of exogenous covariates (explaining ones $=\boldsymbol{X}_f$) with $\forall j \notin I_r, I_f^j=\emptyset$. 
We see that $I_f$ defines the non-null coefficients in $\boldsymbol{\alpha}_j$ (each sub-regression can be very parsimonious).
Then we have an explicit structure $S=(I_f,I_r,p_f,p_r)$ where $p_r=|I_r|$, $\boldsymbol{p}_f=(p_f^1,\dots,p_f^{p_r})$ is the vector of the number of covariates in each sub-regression  and $p_f^j=|I_f^j|$. 
\\
\\

@@

\subsection{A by-product model: marginal regression with decorrelated covariates}
Now we know $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)$ by the structure of sub-regressions, we are able to define a marginal regression model $P(\boldsymbol{Y}|\boldsymbol{X}_f,S)$ based on the reduced set of independent covariates $\boldsymbol{X}_f$ without significant information loss. 
 	\\
Using the partition $\boldsymbol{X}=[\boldsymbol{X}_f,\boldsymbol{X}_r]$ we can rewrite (\ref{regressionsimple}):
	\begin{equation}
			\boldsymbol{Y}_{|\boldsymbol{X}_f,\boldsymbol{X}_r,S}=\boldsymbol{X}_f\boldsymbol{\beta}_f+\boldsymbol{X}_r\boldsymbol{\beta}_r+\boldsymbol{\varepsilon_Y} \label{MainR}
		\end{equation}
		where $\boldsymbol{\beta}=(\boldsymbol{\beta}_f,\boldsymbol{\beta}_r) \in  \mathcal{R}^p$ is the vector of the regression coefficients associated respectively to $\boldsymbol{X}_f$ and $\boldsymbol{I}_n$ the identity matrix. 
We note that (\ref{SR}) and (\ref{MainR}) give also by simple integration on $\boldsymbol{X}_r$ a marginal regression model on $\boldsymbol{Y}$ {\it depending only on uncorrelated covariates $\boldsymbol{X}_f$}:
\begin{eqnarray}
	\boldsymbol{Y}_{|\boldsymbol{X}_f,S}&=&\boldsymbol{X}_f (\boldsymbol{\beta}_f+ \sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j)+  \sum_{j \in I_r}\beta_{j}\boldsymbol{\varepsilon}_j+\boldsymbol{\varepsilon}_Y \label{Trueexpl} \\
	&=&\boldsymbol{X}_f\boldsymbol{\beta}_f^*+\boldsymbol{\varepsilon}_Y^*=\boldsymbol{X}\boldsymbol{\beta}^*+\boldsymbol{\varepsilon}_Y^* \textrm{ where }\boldsymbol{\beta}^*=(\boldsymbol{\beta}_f^*,\boldsymbol{\beta}_r^*) \textrm{ and } \boldsymbol{\beta}_r^*=\boldsymbol{0}\label{modexpl}
\end{eqnarray}
We note that it is simply a linear regression on some of the original covariates so we only made a pretreatment on the dataset by setting some coefficients to $0$ because of correlations. These $0$ won't be interpreted as independence with the response variable but as redundant information due to correlations between the covariates. It is a variable pre-selection independent of the response $\boldsymbol{Y}$.

@@

	 The Mean Squared Error (\textsc{MSE}) on $\hat{\boldsymbol{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\hat{\boldsymbol{\beta}}|\boldsymbol{X})&=&\parallel \operatorname{Bias}\parallel_2^2+\operatorname{Tr}(\operatorname{Var}(\hat{\boldsymbol{\beta}})) \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}_{OLS}|\boldsymbol{X})&=& 0 + \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) %\textrm{ for OLS, and then for the marginal model:}
			 \\
			\textsc{MSE}(\hat{\boldsymbol{\beta}}^*_{OLS}|\boldsymbol{X})&=& \parallel\sum_{j \in I_r}\beta_{j}\boldsymbol{\alpha}_j \parallel_2^2 +\parallel \boldsymbol{\beta}_r\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X}_f' \boldsymbol{X}_f)^{-1})
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $\boldsymbol{X}_f$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $\boldsymbol{x}_3=\boldsymbol{x}_1+\boldsymbol{x}_2+\boldsymbol{\varepsilon}_3$ where $\boldsymbol{\varepsilon}_3\sim{\mathcal{N}(\boldsymbol{0},\sigma_3^2\boldsymbol{I}_n)}$. We also have $\boldsymbol{\beta}=(1,1,1)$ and $\sigma_Y \in \{1,10\}$  . Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg}'s marginal  model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figure \ref{MQE1} shows the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\boldsymbol{\varepsilon)_3}}{\operatorname{Var}(\boldsymbol{x}_3)}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}
